Low-cost linear-scaling \emph{ab initio} molecular dynamics for weakly-interacting systems

### 2017-08-18 almo-langevin.tex

\begin{abstract}
Linear scaling density functional theory based on absolutely localized molecular orbitals is used in conjunction with the Langevin equation of motion to create a low-cost molecular dynamics method for large systems of weakly-interacting molecules. which retains sufficient accuracy to generate correct dynamical properties in water.

@@

Where $m_i$ refers to the mass of the i-th particle, and $r_{i\alpha}$ its position along dimension $\alpha$. 
$f^{\text{SCF}}_{i\alpha}$ is the force calculated using the ALMO SCF method described above for the i-th particle along dimension $\alpha$, $\gamma_L$ is the Langevin scaling factor which sets the strength of both the damping term  $\gamma_L m_i \dot{r}_{i\alpha}$ and the stochastic Gaussian ``white noise'' term $R^{\gamma_L}_{i\alpha} (t)$. 
In order to correctly sample the canonical distribution, the random force must be zero mean. 
As well, the damping and white noise terms must be related by the second fluctuation dissipation theorem.~\cite{a:Kubo-1986,a:langevin-why,b:tuckerman-stat}
%
\begin{align}
\label{eq:stochastic}
\langle R^{\gamma_L}_{i\alpha} (t) \rangle &= 0, \\
\langle R^{\gamma_L}_{i\alpha} (t)  R^{\gamma_L}_{j\beta} (t') \rangle &= 2 k_B T \gamma_L m_i \delta_{ij} \delta_{\alpha\beta} \delta(t-t')
\end{align}
%
This means the damping and white noise terms will work to maintain a constant temperature while also generating the correct Maxwell-Boltzmann distribution of velocities. 
It has previously been shown that Langevin dynamics can be utilized to correct imperfect force calculations while maintaining accurate dynamical properties.~\cite{a:langevin-why,a:2ndcpmd,b:tuckerman-stat,a:ceriotti} 
By utilizing Eq.(\ref{eq:langevin}) in conjunction with the ALMO DFT method of force calculation, we will show that discontinuity in the PES can be ignored for sufficiently large $\gamma_L$. 
Note that increasing $\gamma_L$ increases the sample size required to generate accurate dynamical properties, as random Gaussian noise is introduced.

The concept of balancing an error in force calculation with the Langevin equation can be taken even further, thereby reducing computational overhead even more. 
During the second step of the ALMO SCF method described above, a self-consistent variational approach is taken to converge the non-orthogonal MOs. 
Generally, this step utilizes a iterative search function in order to reduce the absolute norm of the energy gradient to some max allowable value $\epsilon_{SCF}$. 
Previously, $\epsilon_{SCF}$ had to be set very low to ensure convergence of the electron structure calculations. 
By increasing the allowable $\epsilon_{SCF}$, a great reduction in computational cost can be achieved. 
Obviously, this introduces a significant error into the calculation of forces during each time step. 
We assume that this error has the form
%
\begin{align}
\label{eq:assumption}
f^{\text{SCF}}_{i\alpha} = f^{\text{APP}}_{i\alpha} + R^{\Delta}_{i\alpha} (t)
\end{align}
%
Where $f^{\text{SCF}}_{i\alpha}$ is the fully converged ALMO DFT force calculation, $f^{\text{APP}}_{i\alpha}$ is the approximate force calculated by allowing a large $\epsilon_{SCF}$, and $R^{\Delta}_{i\alpha} (t)$ is a stochastic ``white noise'' term that obeys
%
\begin{align}
\label{eq:stochastic2}
\langle R^{\Delta}_{i\alpha} (t) \rangle &= 0, \\
\label{eq:stochastic3}
\langle R^{\Delta}_{i\alpha} (t)  R^{\Delta}_{j\beta} (t') \rangle &= 2 k_B T \Delta m_i \delta_{ij} \delta_{\alpha\beta} \delta(t-t')
\end{align}
%
We will see that this assumption is sufficiently justified later (see fig. \ref{fig:randomforce}). 
The Langevin equation of motion (\ref{eq:langevin}) can be re-written with the two stochastic terms combined into one $\gamma = \gamma_L+\Delta$.
%
\begin{align}
\label{eq:langevin2}
m_i \ddot{r}_{i\alpha} = f^{\text{APP}}_{i\alpha} - \gamma_L m_i \dot{r}_{i\alpha} + R^{\gamma}_{i\alpha} (t)
\end{align}
%
Note that $\Delta$ can be either positive or negative depending on the way in which errors are introduced in the system of interest. 
$\Delta$ was found to be negative for all systems considered here.

\section{Optimizing the values of $\gamma_L$ and $\Delta$}

>>>>

\begin{abstract}
%Linear scaling density functional theory based on absolutely localized molecular orbitals is used in conjunction with the Langevin equation of motion to create a low-cost molecular dynamics method for large systems of weakly-interacting molecules. which retains sufficient accuracy to generate correct dynamical properties in water. 
Linear scaling density functional theory based on absolutely localized molecular orbitals is used in conjunction with the Langevin equation of motion to create a low-cost molecular dynamics method for weakly coupled molecular systems which retains sufficient accuracy to generate accurate dynamical properties in water.

@@

Where $m_i$ refers to the mass of the i-th particle, and $r_{i\alpha}$ its position along dimension $\alpha$. 
$f^{\text{SCF}}_{i\alpha}$ is the force calculated using the ALMO SCF method described above for the i-th particle along dimension $\alpha$, $\gamma_L$ is the Langevin scaling factor which sets the strength of both the damping term  $\gamma_L m_i \dot{r}_{i\alpha}$ and the stochastic Gaussian ``white noise'' term $R^{\gamma_L}_{i\alpha} (t)$. 
In order to correctly sample the canonical distribution, the random force must be zero mean. 
As well, the damping and white noise terms must be related by the second fluctuation dissipation theorem.~\cite{a:Kubo-1986,a:langevin-why,b:tuckerman-stat}
%
\begin{align}
\label{eq:stochastic}
\langle R^{\gamma_L}_{i\alpha} (t) \rangle &= 0, \\
\langle R^{\gamma_L}_{i\alpha} (t)  R^{\gamma_L}_{j\beta} (t') \rangle &= 2 k_B T \gamma_L m_i \delta_{ij} \delta_{\alpha\beta} \delta(t-t')
\end{align}
%
This means the damping and white noise terms will work to maintain a constant temperature while also generating the correct Maxwell-Boltzmann distribution of velocities. 
It has previously been shown that Langevin dynamics can be utilized to correct imperfect force calculations while maintaining accurate dynamical properties.~\cite{a:langevin-why,a:2ndcpmd,b:tuckerman-stat,a:ceriotti} 
By utilizing Eq.(\ref{eq:langevin}) in conjunction with the ALMO DFT method of force calculation, we will show that discontinuity in the PES can be ignored for sufficiently large $\gamma_L$. 
Note that increasing $\gamma_L$ increases the sample size required to generate accurate dynamical properties, as more random Gaussian noise is introduced.

The concept of balancing an error in force calculation with the Langevin equation can be taken even further, thereby reducing computational overhead even more. 
During the second step of the ALMO SCF method described above, a self-consistent variational approach is taken to converge the non-orthogonal MOs. 
Generally, this step utilizes a iterative search function in order to reduce the absolute norm of the energy gradient to some target value $\epsilon_{SCF}$. 
Previously, $\epsilon_{SCF}$ had to be set very low to ensure sufficient convergence of the electron structure calculations. 
By increasing the allowable $\epsilon_{SCF}$, a great reduction in computational cost can be achieved. 
Obviously, this introduces a significant error into the calculation of forces during each time step. 
We assume that this error has the form:
%
\begin{align}
\label{eq:assumption}
f^{\text{SCF}}_{i\alpha} = f^{\text{APP}}_{i\alpha} + R^{\Delta}_{i\alpha} (t)
\end{align}
%
where $f^{\text{SCF}}_{i\alpha}$ is the fully converged ALMO DFT force calculation, $f^{\text{APP}}_{i\alpha}$ is the approximate force calculated by allowing a large $\epsilon_{SCF}$, and $R^{\Delta}_{i\alpha} (t)$ is a stochastic ``white noise'' term that obeys
%
\begin{align}
\label{eq:stochastic2}
\langle R^{\Delta}_{i\alpha} (t) \rangle &= 0, \\
\label{eq:stochastic3}
\langle R^{\Delta}_{i\alpha} (t)  R^{\Delta}_{j\beta} (t') \rangle &= 2 k_B T \Delta m_i \delta_{ij} \delta_{\alpha\beta} \delta(t-t')
\end{align}
%
It's clear from figure \ref{fig:randomforce} that this assumption is sufficiently justified. 
The Langevin equation of motion (\ref{eq:langevin}) can be re-written with the two stochastic terms combined into one $\gamma = \gamma_L+\Delta$.
%
\begin{align}
\label{eq:langevin2}
m_i \ddot{r}_{i\alpha} = f^{\text{APP}}_{i\alpha} - \gamma_L m_i \dot{r}_{i\alpha} + R^{\gamma}_{i\alpha} (t)
\end{align}
%
Note that $\Delta$ can be either positive or negative depending on the way in which errors are introduced in the system of interest. 
$\Delta$ was found to be positive for all systems considered here, increasing the effective random force on the system.

%\section{Optimizing the values of $\gamma_L$ and $\Delta$}
